---
title: "Global Profiling of the Cellular Alternative RNA Splicing Landscape during Virus-Host Interactions: An Analysis"
author: "Emily Mynar & Anna Rees"
date: "`r Sys.Date()`"
output: html_document
---
________________________________________________________________________________
## Data Information
This data was sourced from a study conducted by Boudreault et al, cited below. 
  Alternative splicing (AS) is a method of gene regulation in some Eukaryotes which modifies the sequence of RNA transcripts. AS changes the composition of the proteins resulting from the RNA transcripts through differential choice of exons included in mature mRNAs. This results in higher variability and diversity in the cellular proteome.The study from which this dataset is lifted looks into the global changes in RNA splicing that occurs after infection with a human virus.

Boudreault S, Martenon-Brodeur C, Caron M, Garant JM, Tremblay MP, Armero VE, 
  Durand M, Lapointe E, Thibault P, Tremblay-Létourneau M, Perreault JP, Scott 
  MS, Lemay G, Bisaillon M. Global Profiling of the Cellular Alternative RNA 
  Splicing Landscape during Virus-Host Interactions. PLoS One. 2016 Sep 
  6;11(9):e0161914.doi: 10.1371/journal.pone.0161914. PMID: 27598998; PMCID: PMC5012649.
________________________________________________________________________________ 

## Setup
This data analysis uses additional packages beyond base R, which includes the package manager pacman.

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE,
                      # warning = FALSE,
                      # message = FALSE,
                      fig.align = "center",
                      fig.height = 8)

pacman::p_load(dplyr, ggplot2, DESeq2, tidyverse, forcats, readr, usedist)

theme_set(theme_bw())
```

## Bringing in the metadata_SRP074247.tsv file
This is reading in the metadata into a data frame and doing some preliminary data cleaning, which includes separating values and using the janitor function to clean up the metadata.

```{r read files}

meta <- read.table(file = './Data/SRP074247/metadata_SRP074247.tsv',
           sep = '\t',
           header = TRUE)

janitor::clean_names(meta)
```

## Clean Meta file
It is important to clean the metadata so the data analysis is accurate and streamlined. meta_clean is a cleaned version of the meta data frame. The treatment variables we are interested in are Mock, Reovirus, and Mutant. The names in meta were cleaned to only include the treatment type and the replicate number. First the treatment type and replicate numbers are separated out into their own columns, and then recombined for a cleaner view of what each column is telling us.

```{r clean meta file}
meta_clean <-
  
meta |>
  # Changing columns to choose based on how our dataset is stored
  mutate('sample' = refinebio_accession_code,
         'title' = refinebio_title,
         'treatment' = refinebio_title,
         'replicate' = refinebio_title,
         )

for(i in 1:nrow(meta_clean)){
  if(grepl('Mock', meta_clean$title[i])){
    meta_clean$treatment[i] <- 'Mock'}
  if(grepl('with Reovirus', meta_clean$title[i])){
    meta_clean$treatment[i] <- 'Reovirus'}
  if(grepl('Mutant', meta_clean$title[i])){
    meta_clean$treatment[i] <- 'Mutant'}
}

for(i in 1:nrow(meta_clean)){
  if(grepl('1', meta_clean$title[i])){
    meta_clean$replicate[i] <- '1'}
  if(grepl('2', meta_clean$title[i])){
    meta_clean$replicate[i] <- '2'}
  if(grepl('3', meta_clean$title[i])){
    meta_clean$replicate[i] <- '3'}
}

meta_clean <-
  meta_clean |>
  select(sample, title, treatment, replicate) |>
  janitor::clean_names()

meta_clean
```
## Counts Matrix
// TODO: describe the counts matrix

Reading in the counts matrix
```{r read counts matrix}
# Read in counts matrix
data <- read.table(file = './Data/SRP074247/SRP074247.tsv',
           sep = '\t',
           header = TRUE,
           row.names = 1)

tibble(data)
```
```{r check identical}

identical(colnames(data), meta_clean$sample)

```
This is renaming the columns created above for treatment type and replicate number on the counts matrix

```{r change column names}

colnames(data) <- paste(meta_clean$treatment, meta_clean$replicate, sep = " ")

tibble(data)
```
This is adding the gene name to each row

```{r add gene as column}

data <- rownames_to_column(data, 'gene')

tibble(data)
```
## Pivoting the data frame
Initially, the read matrix had one row for every gene, and each column was the read count for each treatment and replicate. By pivoting the data, we now are able to have more than one row for each gene (there are now nine rows for each gene, as there are nine sample treatment types), and the columns are divided by sample, treatment, replicate, gene, and reads. This way, the data is easier to assess and the data frame contains more information than just the read counts.
```{r pivot data}

data_long <-
  pivot_longer(data = data,
               cols = -gene,
               names_to = 'ID',
               values_to = 'reads') |>
  
  mutate('sample' = ID) |>

  separate(col = 'ID',
           into = c('treatment', 'replicate'),
           sep = ' ') |>
  
  mutate(treatment = as.factor(treatment),
         replicate = as.factor(replicate)) |>
  
  select(sample, treatment, replicate, gene, reads)

tibble(data_long)

```

```{r colors}
colors <- c('Mock' = '#FA8633',
            'Mutant' = '#1C6C7F',
            'Reovirus' = '#9A6DA3')
```
## Checking the quality of the dataset
For gene expression profiling experiments, we hope for between 10-25 million reads per sample. Below is a bar plot showing the sum of the total reads for each sample

```{r check quality}

sum_by_sample <-
  data_long |> 
  group_by(treatment, sample) |> 
  summarize('sum' = sum(reads)) |>
  ungroup()

ggplot(data = data_long,
       mapping = aes(x = sample)) +
          
  geom_col(data = sum_by_sample,
           mapping = aes(y = sum,
                         fill = treatment),
           color = 'black') +
  
  labs(x = 'Sample',
       y = 'Total Reads',
       title = 'Assessing the quality of the dataset using number of reads',
       caption = 'Data from refine.bio\nAccession ID: SRP074247',
       fill = 'Sample type') +
  
  theme(plot.title = element_text(hjust = 0.5,
                                  size = 16),
        axis.text.x = element_text(angle = 90, 
                                   vjust = 0.5, 
                                   hjust = 1),
        axis.title.y = element_text(size = 14),
        axis.title.x = element_text(size = 14),
        legend.position = 'right',
        plot.caption = element_text(face = 'italic')) +
  
  scale_fill_manual(values = colors) +
  
  scale_y_continuous(expand = c(0.01, 0))
```

```{r boxplot}

ggplot(data = sum_by_sample |> group_by(treatment) |> summarize('sum' = sum(sum)),
       mapping = aes(x = sum, 
                     y = treatment,
                     fill = treatment)) +
  
  geom_boxplot(data = data_long |> group_by(reads),
               mapping = aes(x = log2(reads))) +
  
  
  scale_fill_manual(values = colors) +
  
  labs(title = 'Number of reads by sample type',
       x = 'log2(reads)',
       y = NULL,
       fill = 'Sample type') +
  
  theme(plot.title = element_text(hjust = 0.5,
                                  size = 16))
         
         
        
```

```{r t test}

# TODO YOU CANT DO A T TEST WITH MORE THAN TWO LEVELS SO WE WILL COME BACK TO THIS 
```

```{r low abundance histogram}

sum_by_gene <-
  data_long |> 
  group_by(gene) |> 
  summarize('sum' = sum(reads)) |>
  ungroup()

ggplot(data = sum_by_gene,
       mapping = aes(x = log2(sum))) +
  
  geom_histogram(bins = 15,
                 color = 'black',
                 fill = '#1C6C7F') +
  
  labs(x = 'Log-2 transformed counts per gene',
       y = 'Number of genes',
       title = 'Number of reads by gene') +
  
   theme(plot.title = element_text(hjust = 0.5,
                                  size = 16))

```
```{r exclude low abundance}

#high_abundance_genes <-
  #data |>
  #filter_if(is.numeric, all_vars((.) != 0))
  #filter(if_any(-gene, ~ !.x  %in% 0)) # why isn't it getting rid of some of the rows with zero

high_abundance_genes <-
  data_long |>
  pivot_wider(names_from = gene,
              values_from = reads) |>
  select(where(~ all(. != 0)))

high_abundance_long <- 
  high_abundance_genes |>
  pivot_longer(cols = -c(sample, treatment, replicate),
               names_to = 'gene',
               values_to = 'reads')

high_abundance_sum <-
  high_abundance_long |>
  group_by(gene) |> 
  summarize('sum' = sum(reads)) |>
  ungroup()

ggplot(data = high_abundance_sum,
       mapping = aes(x = log2(sum))) +
  
  geom_histogram(bins = 15,
                 color = 'black',
                 fill = '#1C6C7F') +
  
  labs(x = 'Log-2 transformed counts per gene',
       y = 'Number of genes',
       title = 'Number of reads by gene (excluding low abundance genes)') +
  
   theme(plot.title = element_text(hjust = 0.5,
                                  size = 16))


```
```{r distribution per sample}

ggplot(data = high_abundance_long,
       mapping = aes(x = sample)) +
          
  geom_boxplot(mapping = aes(y = log2(reads),
                             fill = treatment),
              color = 'black') +
  
  labs(x = 'Sample ID',
       y = 'Log 2 transformed counts per gene',
       title = 'Number of reads per sample',
       caption = 'Data from refine.bio\nAccession ID: SRP074247',
       fill = 'Sample type') +
  
  theme(plot.title = element_text(hjust = 0.5,
                                  size = 16),
        axis.text.x = element_text(angle = 90, 
                                   vjust = 0.5, 
                                   hjust = 1),
        axis.title.y = element_text(size = 14),
        axis.title.x = element_text(size = 14),
        legend.position = 'right',
        plot.caption = element_text(face = 'italic')) +
  
  scale_fill_manual(values = colors) +
  
  scale_y_continuous(expand = c(0.01, 0))
```

```{r}
sample_names <- c('Mock 1', 'Mock 2', 'Mock 3', 'Mutant 1', 'Mutant 2', 'Mutant 3', 'Reovirus 1', 'Reovirus 2', 'Reovirus 3')
```


```{r sample comparison}
high_abundance_log2 <-
  high_abundance_long |> 
  mutate(reads = log2(reads)) |>
  pivot_wider(names_from = gene,
              values_from = reads) |>
  select(-c(treatment, replicate))

data_dist <- dist(high_abundance_log2, method = "euclidean")
data_dist <- dist_setNames(data_dist, c('Mock 1', 'Mock 2', 'Mock 3', 'Mutant 1', 'Mutant 2', 'Mutant 3', 'Reovirus 1', 'Reovirus 2', 'Reovirus 3'))


plot(hclust(data_dist, method = "complete"), xlab = "Sample", main = NULL)
```

```{r normalizing}

sample_medians <- as.numeric(apply(t(high_abundance_log2), 2, median))

grand_median <- mean(sample_medians)

correction_factors <- grand_median - sample_medians

normalizing <- data.frame(sample_names, correction_factors)

normalizing
```

```{r}

```




words at the end!
