---
title: "Global Profiling of the Cellular Alternative RNA Splicing Landscape during Virus-Host Interactions: An Analysis"
author: "Emily Mynar & Anna Rees"
date: "`r Sys.Date()`"
output: html_document
---
________________________________________________________________________________
## Data Information
This data was sourced from a study conducted by Boudreault et al, cited below. 
  Alternative splicing (AS) is a method of gene regulation in some Eukaryotes which modifies the sequence of RNA transcripts. AS changes the composition of the proteins resulting from the RNA transcripts through differential choice of exons included in mature mRNAs. This results in higher variability and diversity in the cellular proteome.The study from which this dataset is lifted looks into the global changes in RNA splicing that occurs after infection with a human virus.

Boudreault S, Martenon-Brodeur C, Caron M, Garant JM, Tremblay MP, Armero VE, 
  Durand M, Lapointe E, Thibault P, Tremblay-Létourneau M, Perreault JP, Scott 
  MS, Lemay G, Bisaillon M. Global Profiling of the Cellular Alternative RNA 
  Splicing Landscape during Virus-Host Interactions. PLoS One. 2016 Sep 
  6;11(9):e0161914.doi: 10.1371/journal.pone.0161914. PMID: 27598998; PMCID: PMC5012649.
________________________________________________________________________________ 

## Setup
This data analysis uses additional packages beyond base R, which includes the package manager pacman.

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE,
                      # warning = FALSE,
                      # message = FALSE,n
                      fig.align = "center",
                      fig.height = 8)

#install.packages('pacman')
#install.packages(c("BiocManager", "RColorBrewer", "tidyverse", "devtools", "pheatmap", "gprofiler2"))
#install.packages("genekitr")

#BiocManager::install(c("DESeq2", "org.Hs.eg.db", "EnhancedVolcano", "hypeR"))

pacman::p_load(dplyr, ggplot2, DESeq2, tidyverse, forcats, 
               readr, usedist, gridExtra, DESeq2, RColorBrewer, pheatmap,
               genekitr, clusterProfiler, EnhancedVolcano, ggfortify)

theme_set(theme_bw())
```

## Bringing in the metadata_SRP074247.tsv file
This is reading in the metadata into a data frame and doing some preliminary data cleaning, which includes separating values and using the janitor function to clean up the metadata.

#:-)

```{r read files}

meta <- read.table(file = './Data/SRP074247/metadata_SRP074247.tsv',
           sep = '\t',
           header = TRUE)

janitor::clean_names(meta)
```

## Clean Meta file
It is important to clean the metadata so the data analysis is accurate and streamlined. meta_clean is a cleaned version of the meta data frame. The treatment variables we are interested in are Mock, Reovirus, and Mutant. The names in meta were cleaned to only include the treatment type and the replicate number. First the treatment type and replicate numbers are separated out into their own columns, and then recombined for a cleaner view of what each column is telling us.

```{r clean meta file}
meta_clean <-
  
meta |>
  # Changing columns to choose based on how our dataset is stored
  mutate('sample_ID' = refinebio_accession_code,
         'title' = refinebio_title,
         'treatment' = refinebio_title,
         'replicate' = refinebio_title,
         )

for(i in 1:nrow(meta_clean)){
  if(grepl('Mock', meta_clean$title[i])){
    meta_clean$treatment[i] <- 'uninfected'}
  if(grepl('with Reovirus', meta_clean$title[i])){
    meta_clean$treatment[i] <- 'infected'}
  if(grepl('Mutant', meta_clean$title[i])){
    meta_clean$treatment[i] <- 'mutant'}
}

for(i in 1:nrow(meta_clean)){
  if(grepl('1', meta_clean$title[i])){
    meta_clean$replicate[i] <- '1'}
  if(grepl('2', meta_clean$title[i])){
    meta_clean$replicate[i] <- '2'}
  if(grepl('3', meta_clean$title[i])){
    meta_clean$replicate[i] <- '3'}
}

meta_clean <-
  meta_clean |>
  filter(treatment %in% c('uninfected', 'infected')) |>
  mutate('sample' = paste(treatment, replicate, sep = "_")) |>
  select(sample_ID, sample, title, treatment, replicate) |>
  janitor::clean_names()

meta_clean
```
## Counts Matrix
The counts matrix is a data frame which contains the actual data output from the RNA-seq experiment, and will be used for the differential gene expression analysis.
_________________________________________________________________________________

Reading in the counts matrix
```{r read counts matrix}
# Read in counts matrix
data <- read.table(file = './Data/SRP074247/SRP074247.tsv',
           sep = '\t',
           header = TRUE,
           row.names = 1) |>
  select(-c(SRR3471114, SRR3471115, SRR3471116))

head(data)
```
```{r check identical}

identical(colnames(data), meta_clean$sample_id)

```
## Changing the Column Names
This is renaming the columns created above for treatment type (infected or uninfected) and replicate number on the counts matrix

```{r change column names}

colnames(data) <- paste(meta_clean$treatment, meta_clean$replicate, sep = "_")

tibble(data)
```
This is adding the gene name to each row

```{r add gene as column}

data_clean <- rownames_to_column(data, 'gene')

tibble(data_clean)
```
## Pivoting the data frame
Initially, the read matrix had one row for every gene, and each column was the read count for each treatment and replicate. By pivoting the data, we now are able to have more than one row for each gene (there are now nine rows for each gene, as there are nine sample treatment types), and the columns are divided by sample, treatment, replicate, gene, and reads. This way, the data is easier to assess and the data frame contains more information than just the read counts.
```{r pivot data}

data_long <-
  pivot_longer(data = data_clean,
               cols = -gene,
               names_to = 'ID',
               values_to = 'reads') |>
  
  mutate('sample' = ID) |>

  separate(col = 'ID',
           into = c('treatment', 'replicate'),
           sep = '_') |>
  
  mutate(treatment = as.factor(treatment),
         replicate = as.factor(replicate)) |>
  
  select(sample, treatment, replicate, gene, reads)

tibble(data_long)

```

```{r colors}
colors <- c('infected' = '#FA8633',
            'uninfected' = 'lightseagreen')
```

## Checking the quality of the dataset
For gene expression profiling experiments, we hope for between 10-25 million reads per sample. Below is a bar plot showing the sum of the total reads for each sample

We must find if there is a significant different in the total counts by treatment

```{r check quality}

sum_by_sample <-
  data_long |> 
  group_by(treatment, sample) |> 
  summarize('sum' = sum(reads)) |>
  ungroup()

ggplot(data = data_long,
       mapping = aes(x = sample)) +
          
  geom_col(data = sum_by_sample,
           mapping = aes(y = sum,
                         fill = treatment),
           color = 'black') +
  
  labs(x = 'Sample',
       y = 'Total Reads',
       title = 'Assessing the quality of the dataset using number of reads',
       caption = 'Data from refine.bio\nAccession ID: SRP074247',
       fill = 'Sample type') +
  
  theme(plot.title = element_text(hjust = 0.5,
                                  size = 16),
        axis.text.x = element_text(angle = 90, 
                                   vjust = 0.5, 
                                   hjust = 1),
        axis.title.y = element_text(size = 14),
        axis.title.x = element_text(size = 14),
        legend.position = 'right',
        plot.caption = element_text(face = 'italic')) +
  
  scale_fill_manual(values = colors) +
  
  scale_y_continuous(expand = c(0.01, 0))
```

Boxplot of the number of reads by sample type
```{r boxplot}

ggplot(data = sum_by_sample |> group_by(treatment) |> summarize('sum' = sum(sum)),
       mapping = aes(x = sum, 
                     y = treatment,
                     fill = treatment)) +
  
  geom_boxplot(data = data_long |> group_by(reads),
               mapping = aes(x = log2(reads))) +
  
  
  scale_fill_manual(values = colors) +
  
  labs(title = 'Number of reads by sample type',
       x = 'log2(reads)',
       y = NULL,
       fill = 'Sample type') +
  
  theme(plot.title = element_text(hjust = 0.5,
                                  size = 16))
         
         
        
```

## Evalutating statistical significance

We are conducting a T-Test to evaluate whether or not there is a statistical difference in counts by treatment. We are hoping for a high p-value which will indicate no statistical significance in the difference in read counts by treatment.

Rather than conducting a T-test to establish statistical significance, we will use Analysis of variance (ANOVA). This is because we have 3 treatment variables, and a T test is only able to analyze significance between 2 variables.
```{r t test}

t.test(colSums(data) ~ meta_clean$treatment)

```

## Histogram of reads by gene

```{r histogram}


sum_by_gene <-
  data_long |> 
  group_by(gene) |> 
  summarize('sum' = sum(reads)) |>
  ungroup()

ggplot(data = sum_by_gene,
       mapping = aes(x = log2(sum))) +
  
  geom_histogram(bins = 15,
                 color = 'black',
                 fill = '#9A6DA3') +
  
  labs(x = 'Log-2 transformed counts per gene',
       y = 'Number of genes',
       title = 'Number of reads by gene') +
  
   theme(plot.title = element_text(hjust = 0.5,
                                  size = 16))

```

## Exclusing low abundance genes
```{r exclude low abundance}

high_abun_genes <-
  data_long |>
  pivot_wider(names_from = gene,
              values_from = reads) |>
  select(where(~ all(. != 0)))

high_abun_long <- 
  high_abun_genes |>
  pivot_longer(cols = -c(sample, treatment, replicate),
               names_to = 'gene',
               values_to = 'reads')

high_abun_sum <-
  high_abun_long |>
  group_by(gene) |> 
  summarize('sum' = sum(reads)) |>
  ungroup()

ggplot(data = high_abun_sum,
       mapping = aes(x = log2(sum))) +
  
  geom_histogram(bins = 15,
                 color = 'black',
                 fill = '#9A6DA3') +
  
  labs(x = 'Log-2 transformed counts per gene',
       y = 'Number of genes',
       title = 'Number of reads by gene (excluding low abundance genes)') +
  
   theme(plot.title = element_text(hjust = 0.5,
                                  size = 16))


```


```{r distribution per sample}

non_normal_graph <-
  
ggplot(data = high_abun_long,
       mapping = aes(x = sample)) +
          
  geom_boxplot(mapping = aes(y = log2(reads),
                             fill = treatment),
              color = 'black') +
  
  labs(x = 'Sample ID',
       y = 'Log 2 transformed counts per gene',
       title = 'Non-normalized counts',
       caption = 'Data from refine.bio\nAccession ID: SRP074247',
       fill = 'Sample type') +
  
  theme(plot.title = element_text(hjust = 0.5,
                                  size = 16),
        axis.text.x = element_text(angle = 90, 
                                   vjust = 0.5, 
                                   hjust = 1),
        axis.title.y = element_text(size = 14),
        axis.title.x = element_text(size = 14),
        legend.position = 'right',
        plot.caption = element_text(face = 'italic')) +
  
  scale_fill_manual(values = colors) +
  
  scale_y_continuous(expand = c(0.01, 0))

non_normal_graph
```

```{r}
sample_names <- c('uninfected_1', 'uninfected_2', 'uninfected_3', 'infected_1', 'infected_2', 'infected_3')
```


```{r sample comparison}
high_abun_log2 <-
  high_abun_long |> 
  mutate(reads = log2(reads)) |>
  pivot_wider(names_from = gene,
              values_from = reads)

high_abun_log2_flipped <-
  data.frame(t(high_abun_log2 |> select(-c(treatment, replicate)))) |> janitor::row_to_names(row_number = 1) |>
  mutate_all(function(x) as.numeric(as.character(x)))


data_dist <- dist(high_abun_log2, method = "euclidean")
data_dist <- dist_setNames(data_dist, sample_names)


plot(hclust(data_dist, method = "complete"), xlab = "Sample", main = NULL)


```

```{r correction factors}

sample_medians <- (apply(high_abun_log2_flipped, 2, median))

grand_median <- mean(sample_medians)

correction_factors <- grand_median - sample_medians

corrections <- data.frame(correction_factors)

corrections
```

```{r normalizing}

norm_data <- high_abun_log2_flipped

for(i in 1:ncol(norm_data)){
  norm_data[,i] <- high_abun_log2_flipped[,i] + corrections$correction_factors[i]
}

```

```{r pivoting normalized data}

norm_data_long <-
  norm_data |> rownames_to_column('gene') |>
  pivot_longer(cols = -gene,
               names_to = 'ID',
               values_to = 'normalized_reads') |>
  
  mutate('sample' = ID) |>

  separate(col = 'ID',
           into = c('treatment', 'replicate'),
           sep = '_') |>
  
  mutate(treatment = as.factor(treatment),
         replicate = as.factor(replicate)) |>
  
  select(sample, treatment, replicate, gene, normalized_reads)
```

```{r normalized graph}

normal_graph <-
  
ggplot(data = norm_data_long,
       mapping = aes(x = sample)) +
          
  geom_boxplot(mapping = aes(y = normalized_reads,
                             fill = treatment),
              color = 'black') +
  
  labs(x = 'Sample ID',
       y = 'Log 2 transformed counts per gene',
       title = 'Normalized counts',
       caption = 'Data from refine.bio\nAccession ID: SRP074247',
       fill = 'Sample type') +
  
  theme(plot.title = element_text(hjust = 0.5,
                                  size = 16),
        axis.text.x = element_text(angle = 90, 
                                   vjust = 0.5, 
                                   hjust = 1),
        axis.title.y = element_text(size = 14),
        axis.title.x = element_text(size = 14),
        legend.position = 'right',
        plot.caption = element_text(face = 'italic')) +
  
  scale_fill_manual(values = colors) +
  
  scale_y_continuous(expand = c(0.01, 0))

normal_graph
```

```{r both boxplots}

grid.arrange(non_normal_graph, normal_graph, ncol=2)
```

```{r normal tree}
norm_log2 <-
  norm_data_long |>
  mutate(normalized_reads = log2(normalized_reads)) |>
  pivot_wider(names_from = gene,
              values_from = normalized_reads)

norm_dist <- dist(norm_log2, method = "euclidean")
norm_dist <- dist_setNames(norm_dist, sample_names)

plot(hclust(norm_dist, method = "complete"), xlab = "Sample", main = NULL)

```
```{r pca}


PCA <- prcomp(t(norm_dist))

autoplot(PCA, data = data_long, color = 'treatment', main = "Colored by Treatment", size = 3)


```
```{r}

```


```{r create dds}
data_round <- round(data.matrix(data))

#Make sure order is correct
head(data_round,2)
meta_clean[, c('sample')]



dds <- DESeqDataSetFromMatrix(countData = data_round,           
                                       colData = meta_clean,
                                       design = ~ treatment)
dds
```
```{r prefilter dds}
# pre-filter to remove low-read rows
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep,]
```

```{r reference level }
dds$treatment <- relevel(dds$treatment, ref = "uninfected")
```

```{r dds}
# run Differential Expression Analysis
dds <- DESeq(dds)
res <- results(dds)

head(res)
print('Dimensions:')
dim(res)
```

```{r pca 2}

vsd <- vst(dds, blind=FALSE)

plotPCA(vsd, intgroup=c("treatment")) + theme_classic() + scale_color_manual(values = colors)

```

```{r heatmap}
rld <- rlog(dds, blind=FALSE)
rld_mat <- assay(rld)  
rld_cor <- cor(rld_mat) 

heat.colors <- brewer.pal(9, "Spectral")
pheatmap(rld_cor, color = heat.colors,
         breaks = c(0.95, 0.99, 0.995, 0.999, 0.9995, 0.99995))
```

```{r table results}

infected_n_res <- results(dds, contrast = c("treatment", "infected", "uninfected"))
infected_n_sigs <- na.omit(infected_n_res)
infected_n_sigs <- subset(infected_n_sigs, padj < 0.05 & abs(log2FoldChange) > 1)
infected_n_sig_data <- merge(data.frame(infected_n_sigs),
                        data.frame(counts(dds, normalized = TRUE)), 
                        by = "row.names", sort=FALSE)

names(infected_n_sig_data)[1] <- "Ensembl_ID"

infected_n_res_data <- merge(data.frame(infected_n_res),
                        data.frame(counts(dds, normalized = TRUE)),
                        by="row.names", sort=FALSE)

names(infected_n_res_data)[1] <- "Ensembl_ID"

head(res)
```


```{r change gene names}
ids <- infected_n_res_data |> select(Ensembl_ID)

#gene_symbol_map <- transId(id = ids, transTo = "symbol") #default is human! 
#str(infected_n_res_data)
#infected_n_res_data_gene <- left_join(infected_n_res_data, gene_symbol_map,
                     # by=c('Ensembl_ID'='input_id'))

```

```{r}
#output tabular files
write.csv(infected_n_res_data, 
          quote = FALSE,
          file = "infected_vs_uninfected_normalized_matrix.csv")

#infected_n_sig_data <- subset(infected_n_res_data_gene, 
                                 # padj < 0.05 & abs(log2FoldChange) > 1)

write.csv(infected_n_sig_data, 
          quote = FALSE,
          file = "infected_vs_uninfected_deg_padj0.05_log2fc1.csv")
```

```{r}

EnhancedVolcano(infected_n_res_data,
                lab = as.character(infected_n_res_data$Ensembl_ID),
                x = 'log2FoldChange',
                y = 'padj', 
                title = "SLE versus healthy controls",
                subtitle = "Differential expression",
                caption = paste0("Upregulated = 330, Downregulated = 222"),
                xlim = c(-10,10),
                ylim = c(0,20),
                FCcutoff = 1,
                pCutoff = 0.05,
                labSize = 4, 
                axisLabSize = 12, 
                col=c('grey40', 'grey40', 'grey40', '#9A6DA3'),
                legendLabels=c('Not sig.','Log2FC','padj', 'padj & Log2FC'),
                legendPosition = 'right',
                legendLabSize = 10,
                legendIconSize = 5.0,
                gridlines.major = FALSE,
                gridlines.minor = FALSE)
```

```{r}
# plotting upregulated gene

d <- plotCounts(dds, gene="ENSMUSG00000024678", intgroup="treatment", returnData=TRUE)

# Plotting the MOV10 normalized counts, using the samplenames (rownames of d as labels)
ggplot(d, aes(x = treatment, y = count, color = treatment)) + 
  geom_point(position=position_jitter(w = 0.1,h = 0),
             size = 2.5) +
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_color_manual(values = colors) +
  labs(title = "ENSMUSG00000024678",
       x = 'Treatment',
       y = 'Count')

```

```{r}
# plotting downregulated gene
d <- plotCounts(dds, gene="ENSMUSG00000056328", intgroup="treatment", returnData=TRUE)

# Plotting the MOV10 normalized counts, using the samplenames (rownames of d as labels)
ggplot(d, aes(x = treatment, y = count, color = treatment)) + 
  geom_point(position=position_jitter(w = 0.1,h = 0),
             size = 2.5) +
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_color_manual(values = colors) +
  labs(title = "ENSMUSG00000056328",
       x = 'Treatment',
       y = 'Count')

```

```{r}
## Order results by padj values
top20_sigOE_genes <- infected_n_res_data |>
        arrange(padj) |> 	#Arrange rows by padj values
        pull(Ensembl_ID) |>		#Extract character vector of ordered genes
        head(n=20) 		#Extract the first 20 genes

## normalized counts for top 20 significant genes
top20_sigOE_norm <- infected_n_sig_data |>
        filter(Ensembl_ID %in% top20_sigOE_genes)

# Gathering the columns to have normalized counts to a single column
gathered_top20_sigOE <- top20_sigOE_norm |>
  gather(colnames(top20_sigOE_norm)[8:13], key = "sample", value = "normalized_counts")

## check the column header in the "gathered" data frame
head(gathered_top20_sigOE)

gathered_top20_sigOE <- inner_join(meta_clean, gathered_top20_sigOE)
```
```{r}
## plot using ggplot2

ggplot(data = gathered_top20_sigOE,
       mapping = aes(x = as.character(Ensembl_ID), y = normalized_counts, color = treatment)) +
  geom_point() +
  scale_y_log10() +
  xlab("Gene") +
  ylab("log10 Normalized Counts") +
  ggtitle("Top 20 Significant DE Genes") +
  theme_bw() +
  scale_color_manual(values = colors) +
      
theme(axis.text.x = element_text(angle = , hjust = 1)) +
theme(plot.title = element_text(hjust = 0.5))
```

