---
title: "Global Profiling of the Cellular Alternative RNA Splicing Landscape during Virus-Host Interactions: An Analysis"
author: "Emily Mynar & Anna Rees"
date: "`r Sys.Date()`"
output: html_document
---
________________________________________________________________________________
## Data Information
This data was sourced from a study conducted by Boudreault et al, cited below. 
  Alternative splicing (AS) is a method of gene regulation in some Eukaryotes which modifies the sequence of RNA transcripts. Alternative Splicing changes the composition of the proteins resulting from the RNA transcripts through differential choice of exons included in mature mRNAs. This results in higher variability and diversity in the cellular proteome.The study from which this dataset is lifted looks into the global changes in RNA splicing that occurs after infection with a human virus.

<ins>**Study Citation**</ins>:
Boudreault S, Martenon-Brodeur C, Caron M, Garant JM, Tremblay MP, Armero VE, 
  Durand M, Lapointe E, Thibault P, Tremblay-Létourneau M, Perreault JP, Scott 
  MS, Lemay G, Bisaillon M. Global Profiling of the Cellular Alternative RNA 
  Splicing Landscape during Virus-Host Interactions. PLoS One. 2016 Sep 
  6;11(9):e0161914.doi: 10.1371/journal.pone.0161914. PMID: 27598998; PMCID: PMC5012649.
________________________________________________________________________________ 

## Setup
This data analysis uses additional packages beyond base R, which includes the package manager pacman.

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE,
                      #warning = FALSE,
                      #message = FALSE,
                      fig.align = "center",
                      fig.height = 8)


if(!require('pacman')){install.packages("pacman", repos = "http://cran.us.r-project.org")}

pacman::p_load(dplyr, ggplot2, DESeq2, tidyverse, forcats, 
               readr, usedist, gridExtra, DESeq2, RColorBrewer, pheatmap,
               genekitr, clusterProfiler, EnhancedVolcano, ggfortify, biomartr,
               hypeR, gprofiler2, BiocManager, DESeq2, org.Hs.eg.db, EnhancedVolcano,
               biomaRt)

#BiocManager::install(c("DESeq2", "org.Hs.eg.db", "EnhancedVolcano", "hypeR", "biomaRt"))


theme_set(theme_bw())
```

## Bringing in the metadata_SRP074247.tsv file
This is reading in the metadata into a data frame and doing some preliminary data cleaning, which includes separating values and using the janitor function to clean up the metadata.


```{r read files}

meta <- read.table(file = './Data/SRP074247/metadata_SRP074247.tsv',
           sep = '\t',
           header = TRUE)

janitor::clean_names(meta)
```

## Clean Meta file
It is important to clean the metadata so the data analysis is accurate and streamlined. meta_clean is a cleaned version of the meta data frame. The treatment variables we are interested in are Mock, Reovirus, and Mutant. The names in meta were cleaned to only include the treatment type and the replicate number. First the treatment type and replicate numbers are separated out into their own columns, and then recombined for a cleaner view of what each column is telling us.

```{r clean meta file}
meta_clean <-
  
meta |>
  # Changing columns to choose based on how our dataset is stored
  mutate('sample_ID' = refinebio_accession_code,
         'title' = refinebio_title,
         'treatment' = refinebio_title,
         'replicate' = refinebio_title,
         )

for(i in 1:nrow(meta_clean)){
  if(grepl('Mock', meta_clean$title[i])){
    meta_clean$treatment[i] <- 'uninfected'}
  if(grepl('with Reovirus', meta_clean$title[i])){
    meta_clean$treatment[i] <- 'infected'}
  if(grepl('Mutant', meta_clean$title[i])){
    meta_clean$treatment[i] <- 'mutant'}
}

for(i in 1:nrow(meta_clean)){
  if(grepl('1', meta_clean$title[i])){
    meta_clean$replicate[i] <- '1'}
  if(grepl('2', meta_clean$title[i])){
    meta_clean$replicate[i] <- '2'}
  if(grepl('3', meta_clean$title[i])){
    meta_clean$replicate[i] <- '3'}
}

meta_clean <-
  meta_clean |>
  filter(treatment %in% c('uninfected', 'infected')) |>
  mutate('sample' = paste(treatment, replicate, sep = "_")) |>
  dplyr::select(sample_ID, sample, title, treatment, replicate) |>
  janitor::clean_names()

meta_clean
```
## Counts Matrix
The counts matrix is a data frame which contains the actual data output from the RNA-seq experiment, and will be used for the differential gene expression analysis.

Reading in the counts matrix
```{r read counts matrix}
# Read in counts matrix
data <- read.table(file = './Data/SRP074247/SRP074247.tsv',
           sep = '\t',
           header = TRUE,
           row.names = 1) |>
  dplyr::select(-c(SRR3471114, SRR3471115, SRR3471116))

head(data)
```
```{r check identical}

identical(colnames(data), meta_clean$sample_id)

```
This boolean flag checks that the cleaned data matches the original data, which it does.

## Changing the Column Names
This is renaming the columns created above for treatment type (infected or uninfected) and replicate number on the counts matrix

```{r change column names}

colnames(data) <- paste(meta_clean$treatment, meta_clean$replicate, sep = "_")

tibble(data)
```
This is adding the gene name to each row

```{r add gene as column}

data_clean <- rownames_to_column(data, 'gene')

tibble(data_clean)
```
## Pivoting the data frame
Initially, the read matrix had one row for every gene, and each column was the read count for each treatment and replicate. By pivoting the data, we now are able to have more than one row for each gene (there are now nine rows for each gene, as there are nine sample treatment types), and the columns are divided by sample, treatment, replicate, gene, and reads. This way, the data is easier to assess and the data frame contains more information than just the read counts.
```{r pivot data}

data_long <-
  pivot_longer(data = data_clean,
               cols = -gene,
               names_to = 'ID',
               values_to = 'reads') |>
  
  mutate('sample' = ID) |>

  separate(col = 'ID',
           into = c('treatment', 'replicate'),
           sep = '_') |>
  
  mutate(treatment = as.factor(treatment),
         replicate = as.factor(replicate)) |>
  
  dplyr::select(sample, treatment, replicate, gene, reads)

tibble(data_long)

```

```{r colors}
colors <- c('infected' = '#FA8633',
            'uninfected' = 'lightseagreen')
```

## Checking the quality of the dataset
For gene expression profiling experiments, we hope for between 10-25 million reads per sample. Below is a bar plot showing the sum of the total reads for each sample

We must find if there is a significant different in the total counts by treatment

```{r check quality}

sum_by_sample <-
  data_long |> 
  group_by(treatment, sample) |> 
  summarize('sum' = sum(reads)) |>
  ungroup()

ggplot(data = data_long,
       mapping = aes(x = sample)) +
          
  geom_col(data = sum_by_sample,
           mapping = aes(y = sum,
                         fill = treatment),
           color = 'black') +
  
  labs(x = 'Sample',
       y = 'Total Reads',
       title = 'Assessing the quality of the dataset using number of reads',
       caption = 'Data from refine.bio\nAccession ID: SRP074247',
       fill = 'Sample type') +
  
  theme(plot.title = element_text(hjust = 0.5,
                                  size = 16),
        axis.text.x = element_text(angle = 90, 
                                   vjust = 0.5, 
                                   hjust = 1),
        axis.title.y = element_text(size = 14),
        axis.title.x = element_text(size = 14),
        legend.position = 'right',
        plot.caption = element_text(face = 'italic')) +
  
  scale_fill_manual(values = colors) +
  
  scale_y_continuous(expand = c(0.01, 0))
```
**Figure 1. A barplot assessing the quality of the dataset used.** For each gene expression profiling experiment, a good quality dataset will have 10-25 million reads per sample. As shown in the plot, all six samples are within the range desired.


```{r boxplot}

ggplot(data = sum_by_sample |> group_by(treatment) |> summarize('sum' = sum(sum)),
       mapping = aes(x = sum, 
                     y = treatment,
                     fill = treatment)) +
  
  geom_boxplot(data = data_long |> group_by(reads),
               mapping = aes(x = log2(reads))) +
  
  
  scale_fill_manual(values = colors) +
  
  labs(title = 'Number of reads by sample type',
       x = 'log2(reads)',
       y = NULL,
       fill = 'Sample type') +
  
  theme(plot.title = element_text(hjust = 0.5,
                                  size = 16))
         
         
        
```
**Figure 2. A Boxplot of the number of reads by sample type.** It is important to analyze the read counts between infected vs uninfected because if there are statistical differences, this will impact downstream biological interpretation.

## Evalutating statistical significance

We are conducting a T-Test to evaluate whether or not there is a statistical difference in counts by treatment. We are hoping for a high p-value which will indicate no statistical significance in the difference in read counts by treatment.

Rather than conducting a T-test to establish statistical significance, we will use Analysis of variance (ANOVA). This is because we have 3 treatment variables, and a T test is only able to analyze significance between 2 variables.
```{r t test}

t.test(colSums(data) ~ meta_clean$treatment)

```
The p-value is .2935, so there is **not** a significant difference between total number of counts by treatment group.

## Histogram of reads by gene

```{r histogram}


sum_by_gene <-
  data_long |> 
  group_by(gene) |> 
  summarize('sum' = sum(reads)) |>
  ungroup()

ggplot(data = sum_by_gene,
       mapping = aes(x = log2(sum))) +
  
  geom_histogram(bins = 15,
                 color = 'black',
                 fill = '#9A6DA3') +
  
  labs(x = 'Log-2 transformed counts per gene',
       y = 'Number of genes',
       title = 'Number of reads by gene') +
  
   theme(plot.title = element_text(hjust = 0.5,
                                  size = 16))

```
**Figure 3. A histogram of the sum number of reads across each gene.** This histogram is including low-abundance gene. There are six samples worth of read counts included in this histogram. 

## Excluding low abundance genes
```{r exclude low abundance}

high_abun_genes <-
  data_long |>
  pivot_wider(names_from = gene,
              values_from = reads) |>
  dplyr::select(where(~ all(. != 0)))

high_abun_long <- 
  high_abun_genes |>
  pivot_longer(cols = -c(sample, treatment, replicate),
               names_to = 'gene',
               values_to = 'reads')

high_abun_sum <-
  high_abun_long |>
  group_by(gene) |> 
  summarize('sum' = sum(reads)) |>
  ungroup()

ggplot(data = high_abun_sum,
       mapping = aes(x = log2(sum))) +
  
  geom_histogram(bins = 15,
                 color = 'black',
                 fill = '#9A6DA3') +
  
  labs(x = 'Log-2 transformed counts per gene',
       y = 'Number of genes',
       title = 'Number of reads by gene (excluding low abundance genes)') +
  
   theme(plot.title = element_text(hjust = 0.5,
                                  size = 16))


```
**Figure 4. A histogram of the sum number of reads across each gene.** This histogram is excluding low abundance genes, or genes with read counts of zero. There are six samples worth of read counts included in this histogram. 

```{r distribution per sample}

non_normal_graph <-
  
ggplot(data = high_abun_long,
       mapping = aes(x = sample)) +
          
  geom_boxplot(mapping = aes(y = log2(reads),
                             fill = treatment),
              color = 'black') +
  
  labs(x = 'Sample ID',
       y = 'Log 2 transformed counts per gene',
       title = 'Non-normalized counts',
       caption = 'Data from refine.bio\nAccession ID: SRP074247',
       fill = 'Sample type') +
  
  theme(plot.title = element_text(hjust = 0.5,
                                  size = 16),
        axis.text.x = element_text(angle = 90, 
                                   vjust = 0.5, 
                                   hjust = 1),
        axis.title.y = element_text(size = 14),
        axis.title.x = element_text(size = 14),
        legend.position = 'right',
        plot.caption = element_text(face = 'italic')) +
  
  scale_fill_manual(values = colors) +
  
  scale_y_continuous(expand = c(0.01, 0))

non_normal_graph
```
**Figure 5. A non-normalized boxplot of the sum number of reads across each gene.** The boxplot provides insight into the distribution of read counts per sample. 

```{r}
sample_names <- c('uninfected_1', 'uninfected_2', 'uninfected_3', 'infected_1', 'infected_2', 'infected_3')
```


```{r sample comparison}
high_abun_log2 <-
  high_abun_long |> 
  mutate(reads = log2(reads)) |>
  pivot_wider(names_from = gene,
              values_from = reads)

high_abun_log2_flipped <-
  data.frame(t(high_abun_log2 |> dplyr::select(-c(treatment, replicate)))) |> janitor::row_to_names(row_number = 1) |>
  mutate_all(function(x) as.numeric(as.character(x)))


data_dist <- dist(high_abun_log2, method = "euclidean")
data_dist <- dist_setNames(data_dist, sample_names)


plot(hclust(data_dist, method = "complete"), xlab = "Sample", main = NULL)


```
**Figure 6. A dendrogram comparing the similarity of read count data across the different samples based on non-normalized data. ** The three infected samples are most similar to one another, and the three uninfected samples are most similar to one another, which is to be expected.

## Normalization

Generating correction factors for the data where the median of each row is subtracted from the mean of the row. We then apply this correction factor to each sample.
```{r correction factors}

sample_medians <- (apply(high_abun_log2_flipped, 2, median))

grand_median <- mean(sample_medians)

correction_factors <- grand_median - sample_medians

corrections <- data.frame(correction_factors)

corrections
```

```{r normalizing}

norm_data <- high_abun_log2_flipped

for(i in 1:ncol(norm_data)){
  norm_data[,i] <- high_abun_log2_flipped[,i] + corrections$correction_factors[i]
}

```

```{r pivoting normalized data}

norm_data_long <-
  norm_data |> rownames_to_column('gene') |>
  pivot_longer(cols = -gene,
               names_to = 'ID',
               values_to = 'normalized_reads') |>
  
  mutate('sample' = ID) |>

  separate(col = 'ID',
           into = c('treatment', 'replicate'),
           sep = '_') |>
  
  mutate(treatment = as.factor(treatment),
         replicate = as.factor(replicate)) |>
  
  dplyr::select(sample, treatment, replicate, gene, normalized_reads)
```

```{r normalized graph}

normal_graph <-
  
ggplot(data = norm_data_long,
       mapping = aes(x = sample)) +
          
  geom_boxplot(mapping = aes(y = normalized_reads,
                             fill = treatment),
              color = 'black') +
  
  labs(x = 'Sample ID',
       y = 'Log 2 transformed counts per gene',
       title = 'Normalized counts',
       caption = 'Data from refine.bio\nAccession ID: SRP074247',
       fill = 'Sample type') +
  
  theme(plot.title = element_text(hjust = 0.5,
                                  size = 16),
        axis.text.x = element_text(angle = 90, 
                                   vjust = 0.5, 
                                   hjust = 1),
        axis.title.y = element_text(size = 14),
        axis.title.x = element_text(size = 14),
        legend.position = 'right',
        plot.caption = element_text(face = 'italic')) +
  
  scale_fill_manual(values = colors) +
  
  scale_y_continuous(expand = c(0.01, 0))

normal_graph
```
**Figure 7.  A normalized boxplot of the sum number of reads across each gene.** The boxplot provides insight into the distribution of read counts per sample. 
 

```{r both boxplots}

grid.arrange(non_normal_graph, normal_graph, ncol=2)
```
**Figure 8. A side by side comparison of the non-normalized vs normalized counts boxplots, to see what difference the normalization made in the figure outcome. ** 


```{r normal tree}
norm_log2 <-
  norm_data_long |>
  mutate(normalized_reads = log2(normalized_reads)) |>
  pivot_wider(names_from = gene,
              values_from = normalized_reads)

norm_dist <- dist(norm_log2, method = "euclidean")
norm_dist <- dist_setNames(norm_dist, sample_names)

plot(hclust(norm_dist, method = "complete"), xlab = "Sample", main = NULL)

```
**Figure 9. A dendrogram comparing the similarity of read count data across the different samples based on normalized data. ** The three infected samples are most similar to one another, and the three uninfected samples are most similar to one another, which is to be expected.

## Principal Component Analysis
Principle Component Analysis is a technique for analyzing large datasets with multiple features to be considered. By using PCA, one is able to maximize interpretability while maintaining the maximum amount of information. T

```{r pca}


PCA <- prcomp(t(norm_dist))

autoplot(PCA, data = data_long, color = 'treatment', main = "Colored by Treatment", size = 3)


```
**Figure 10. PCA which colors by treatment.** This PCA1 gives a percentage of variance of 42.59%, which indicates 42.59% of the variance in the data can be explained based on this variable, and PCA2 gives a percentage of variance of 19.93%. These variables are the two strongest sources of variance in the sample set.

## Creating a DESeq Object
```{r create dds}
data_round <- 
  round(data.matrix(data))

#Make sure order is correct
head(data_round,2)
meta_clean[, c('sample')]



dds <- DESeqDataSetFromMatrix(countData = data_round,           
                                       colData = meta_clean,
                                       design = ~ treatment)


```

## Pre-filtering DDS
By pre-filtering the DESeq2 functions, we are able to eliminate rows with very few reads and therefore reduce the memory used by the dds object. This improves speed and visualization quality.
```{r prefilter dds}
# pre-filter to remove low-read rows
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep,]
```

We must choose a reference level for factors based on treatment type.
```{r reference level }
dds$treatment <- relevel(dds$treatment, ref = "uninfected")
```

## Running Differential Expression Analysis
```{r dds}
# run Differential Expression Analysis
dds <- DESeq(dds)
res <- results(dds)

head(res)
print('Dimensions:')
dim(res)
```

```{r pca 2}

vsd <- vst(dds, blind=FALSE)

plotPCA(vsd, intgroup=c("treatment")) + theme_classic() + scale_color_manual(values = colors)

```
**Figure 11. Principal Component Analysis using variance stabilizing transformation (vst).** This data is transformed on a log2 scale and then normalized with respect of library size. The transformation is <ins>not blind</ins> because the experiment design has a large difference in counts between the treatment types, and blind dispersion estimates wil, overly shrink the values towards one another. This plot demonstrates a PCA based on the first two principle components of the data.

## Hierarchical Clustering Heatmap
```{r heatmap}
rld <- rlog(dds, blind=FALSE)
rld_mat <- assay(rld)  
rld_cor <- cor(rld_mat) 

heat.colors <- brewer.pal(6, "Blues")
pheatmap(rld_cor, color = heat.colors,
         breaks = c(0.97, 0.99, 0.995, 0.999, 0.9995, 0.99995))
```
**Figure 12. A hierarchical clustering heatmap of the similarities between the read counts in each sample.** This heatmap is comparing the similarity of each sample to each other sample, and the darker blue the sample is, the more similar the two are to one another. The uninfected samples are most similar to one another, with uninfected_1 and uninfected_2 being the most similar to each other. The infected samples are each similarly different to one another, which is indicated in the lightest blue shade.

## Tabular results
```{r table results}

infected_n_res <- results(dds, contrast = c("treatment", "infected", "uninfected"))
infected_n_sigs <- na.omit(infected_n_res)
infected_n_sigs <- subset(infected_n_sigs, padj < 0.05 & abs(log2FoldChange) > 1)
infected_n_sig_data <- merge(data.frame(infected_n_sigs),
                        data.frame(counts(dds, normalized = TRUE)), 
                        by = "row.names", sort=FALSE)

names(infected_n_sig_data)[1] <- "Ensembl_ID"

infected_n_res_data <- merge(data.frame(infected_n_res),
                        data.frame(counts(dds, normalized = TRUE)),
                        by="row.names", sort=FALSE)

names(infected_n_res_data)[1] <- "Ensembl_ID"  

head(res)
```

## Gene ID Conversion
We are converting gene IDs from Ensemble ID to symbols for a clearer picture
```{r get gene symbol name}

ids <- as.character(infected_n_res_data$Ensembl_ID)

gene_list <- biomaRt::getBM(filter = 'ensembl_gene_id',
                  attributes = c("ensembl_gene_id","uniprot_gn_symbol"), 
                  values = ids,
                  mart = biomaRt::useMart(biomart="ensembl", dataset="mmusculus_gene_ensembl"))

```

Adding the new symbols to the dataset
```{r add symbol to datasets}

infected_n_res_data <- merge(infected_n_res_data, gene_list, by.x="Ensembl_ID", by.y="ensembl_gene_id")

infected_n_sig_data <- merge(infected_n_sig_data, gene_list, by.x="Ensembl_ID", by.y="ensembl_gene_id")

```


```{r}
#output tabular files
write.csv(infected_n_res_data, 
          quote = FALSE,
          file = "infected_vs_uninfected_normalized_matrix.csv")

#infected_n_sig_data <- subset(infected_n_res_data_gene, 
                                 # padj < 0.05 & abs(log2FoldChange) > 1)

write.csv(infected_n_sig_data, 
          quote = FALSE,
          file = "infected_vs_uninfected_deg_padj0.05_log2fc1.csv")
```

## Volcano Plot
```{r}
EnhancedVolcano(infected_n_res_data,
                lab = as.character(infected_n_res_data$uniprot_gn_symbol),
                x = 'log2FoldChange',
                y = 'padj', 
                title = "Infected Versus Uninfected Controls",
                subtitle = "Differential expression",
                caption = paste0("Upregulated = 330, Downregulated = 222"),
                xlim = c(-10,10),
                ylim = c(0,20),
                FCcutoff = 1,
                pCutoff = 0.05,
                labSize = 4, 
                axisLabSize = 12, 
                col=c('grey40', 'grey40', 'grey40', '#FA8633'),
                legendLabels=c('Not sig.','Log2FC','padj', 'padj & Log2FC'),
                legendPosition = 'right',
                legendLabSize = 10,
                legendIconSize = 5.0,
                gridlines.major = FALSE,
                gridlines.minor = FALSE)
```
**Figure 13. A Volcano plot of the infected samples vs uninfected samples.** Log transformed adjusted p-values are plotted on the y-axis while the log2 fold change is plotted on the x-axis.  

```{r}
# plotting upregulated gene

d <- plotCounts(dds, gene="ENSMUSG00000024678", intgroup="treatment", returnData=TRUE)

# Plotting the MOV10 normalized counts, using the samplenames (rownames of d as labels)
ggplot(d, aes(x = treatment, y = count, color = treatment)) + 
  geom_point(position=position_jitter(w = 0.1,h = 0),
             size = 2.5) +
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_color_manual(values = colors) +
  labs(title = "Ms4a4d",
       x = 'Treatment',
       y = 'Count')

```
**Figure 14. Demonstration of up-regulation using Ms4a4d gene.** There is much higher gene expression in the infected samples compared to the uninfected sample, indicating this gene's expression is impacted by virus-host interactions.
```{r}
# plotting downregulated gene
d <- plotCounts(dds, gene="ENSMUSG00000056328", intgroup="treatment", returnData=TRUE)

# Plotting the MOV10 normalized counts, using the samplenames (rownames of d as labels)
ggplot(d, aes(x = treatment, y = count, color = treatment)) + 
  geom_point(position=position_jitter(w = 0.1,h = 0),
             size = 2.5) +
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_color_manual(values = colors) +
  labs(title = "Myh1",
       x = 'Treatment',
       y = 'Count')

```
**Figure 14. Demonstration of down-regulation using Myh1 gene.** There is much higher gene expression in the uninfected samples compared to the uninfected sample, indicating this gene's expression is impacted by virus-host interactions.

## Pathway Analysis

Gathering the top 20 differentially expressed genes in the dataset.
```{r}
## Order results by padj values
top20_sigOE_genes <- infected_n_res_data |>
        arrange(padj) |> 	#Arrange rows by padj values
        pull(uniprot_gn_symbol) |>		#Extract character vector of ordered genes
        head(n=20) 		#Extract the first 20 genes

## normalized counts for top 20 significant genes
top20_sigOE_norm <- infected_n_sig_data |>
        filter(uniprot_gn_symbol %in% top20_sigOE_genes)

# Gathering the columns to have normalized counts to a single column
gathered_top20_sigOE <- top20_sigOE_norm |>
  gather(colnames(top20_sigOE_norm)[8:13], key = "sample", value = "normalized_counts")

## check the column header in the "gathered" data frame
head(gathered_top20_sigOE)

gathered_top20_sigOE <- inner_join(meta_clean, gathered_top20_sigOE)
```

```{r}
## plot using ggplot2

ggplot(data = gathered_top20_sigOE,
       mapping = aes(x = as.character(uniprot_gn_symbol), y = normalized_counts, color = treatment)) +
  geom_point() +
  scale_y_log10() +
  xlab("Gene") +
  ylab("log10 Normalized Counts") +
  ggtitle("Top 20 Significant Differentially Expressed Genes") +
  theme_bw() +
  scale_color_manual(values = colors) +
      
theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
theme(plot.title = element_text(hjust = 0.5))
```
**Figure 15. Plot of the normalized read counts of the top 20 most significant differentially expressed genes in the dataset.**  

To determine whether certain items are over or underrepresented, we are using a hypergeometric test to determine the probability of having our observed proportion of genes. The KEGG database is used to compare the data against a manually curated database. 
```{r}
KEGG <- msigdb_gsets(species="Mus musculus", 
                     category="C2", 
                     subcategory="CP:KEGG",
                     clean = TRUE)
```

```{r}
hyp_kegg <- hypeR(gene_list$uniprot_gn_symbol, 
                  KEGG, 
                  test="hypergeometric", 
                  background=50000, 
                  fdr=0.05, 
                  plotting=TRUE)
```

```{r}
hyp_dots(hyp_kegg, 
         title="KEGG", 
         abrv=30, 
         val = "fdr")
```
**Figure 16. Dot plot of the top enriched genesets from the KEGG database.** The color of the dot is indicative of its significance, and its size is indicative of its geneset size.


## Hallmark Pathways
Hallmark set summarized the information across multiple genesets, and reduced redundancy.
```{r}
HALLMARK <- msigdb_gsets("Mus musculus", 
                         "H", 
                         "",
                         clean = TRUE)

# Hallmark
hyp_Hallmark <- hypeR(gene_list$uniprot_gn_symbol, 
                      HALLMARK, 
                      test="hypergeometric", 
                      background=50000, 
                      fdr=0.05, 
                      plotting=TRUE)

#show top pathway
hyp_dots(hyp_Hallmark, 
         title="HALLMARK", 
         abrv=50, 
         val = "fdr")
```
**Figure 17. Dot Plot based on the Hallmark database.** The color is indicative of the geneset significance, and its size is indicative of the geneset size. 


## Gene Ontology (GO)
```{r}
GOBP<- msigdb_gsets(species = "Mus musculus", 
                    "C5", 
                    "GO:BP",
                    clean = TRUE)
```

```{r}
# GOBP
hyp_GOBP <- hypeR(gene_list$uniprot_gn_symbol, 
                  GOBP, 
                  test="hypergeometric", 
                  background=50000, 
                  fdr=0.05, 
                  plotting=TRUE)

#show top pathway
hyp_dots(hyp_GOBP, title="GO: Biological Pathways", abrv=50, val = "fdr")
```
**Figure 18. Gene Ontology biological pathway visualization.** This is referring to the biological role involving each gene. The color insicates the significance of the biologicak pathway, and the size of the dot indicates the geneset size. 


## Separating Up- and Down- regulated DE Genes
```{r}
up_degs <- subset(infected_n_sig_data, log2FoldChange > 0)
up_degs <- up_degs$uniprot_gn_symbol
up_degs <- na.omit(up_degs)


down_degs <- subset(infected_n_sig_data, log2FoldChange < 0)
down_degs <- down_degs$uniprot_gn_symbol
down_degs <- na.omit(down_degs)
```

```{r}
hyp_KEGG_up <- hypeR(up_degs,
                     KEGG, 
                     test="hypergeometric", 
                     background=50000, 
                     fdr=0.05, 
                     plotting=TRUE)

#show top pathway
#plot_KEGG_up <- 
  hyp_dots(hyp_KEGG_up, title="KEGG Pathways Upregulated in infected samples", abrv=50, val = "fdr")

#ggsave("Dotplot_KEGG_upregulated.png")
```
**Figure 19. Pathway analysis using only upregulated genes.**

```{r}
# KEGG_down
hyp_KEGG_down <- hypeR(down_degs, 
                       KEGG, 
                       test="hypergeometric",
                       background=50000, 
                       fdr=0.05, 
                       plotting=TRUE)

#plot_KEGG_down <- 
hyp_dots(hyp_KEGG_down, title="KEGG Pathways Downregulated in infected samples", abrv=50, val = "fdr")

#ggsave("Dotplot_KEGG_downregulated.png")
```
**Figure 20. Pathway analysis using only downregulated genes.**

```{r}
sle_vs_n_gost <- gost(query = gene_list$uniprot_gn_symbol, 
                      organism = "mmusculus", 
                      sources = c("GO:BP", "KEGG", "REAC", "WP"),
                      significant = TRUE, 
                      correction_method ="fdr",
                      domain_scope="annotated")

gostplot(sle_vs_n_gost, interactive = TRUE)
```
**Figure 21. Gostplot Visualization of many pathways databases with gprofiler.** This is a statistical enrichment analysis which finds over-representation of information from various networks at the same time. The vector of sources used includes GO:BP, KEGG, REAC, and WP. 

